---
title: "Spatio-temporal Models for Finite Population Prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatio-temporal Models for Finite Population Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```

In addition to the `FPSpatioTemp` package, we also use the `ggplot2`, `dplyr`, and `tibble` packages in the tidyverse:

```{r}
library(FPSpatioTemp)
library(ggplot2)
library(dplyr)
library(tibble)
```

## Toy Example: Data Set-Up

In order to use the functions in the package, the data frame needs to have a few characteristics. At minimum, the data frame needs four variables: a variable for the x-coordinates, a variable for the y-coordinates, a variable for the "time"-coordinate, and a variable for the response. The variable for the time coordinate should either be a `<dbl>` variable or an `<int>` variable; the time should not be of class `<date>`. The response variable should be `NA` for any spatial locations that were not sampled at a particular time point. For example, a toy data set with 4 spatial locations and 2 time points might look like:

```{r}
toy_df <- tibble::tibble(xcoordvar = c(1, 1, 2, 2, 1, 1, 2, 2),
                         ycoordvar = c(1, 2, 1, 2, 1, 2, 1, 2),
                         tcoordvar = c(1, 1, 1, 1, 2, 2, 2, 2),
                         response = c(4, 2, NA, 1, 0, 0, NA, 5)
)
toy_df
```

The data does not need to be ordered in any particular way. We anticipate that, in some contexts, data might only contain non-missing values for the response, as below:

```{r}
toy_df2 <- tibble::tibble(xcoordvar = c(1, 1, 2, 1, 1, 2),
                         ycoordvar = c(1, 2, 2, 1, 2, 2),
                         tcoordvar = c(1, 1, 1, 2, 2, 2),
                         response = c(4, 2, 1, 0, 0, 5)
)
```

To add missing response values for combinations of an x-coordinate, y-coordinate, and time-coordinate variables that are not in the data, we can use the `complete()` function from the `tidyr` package:

```{r}
toy_df2 |> tidyr::complete(xcoordvar, ycoordvar, tcoordvar)
```

For the rest of the vignette, we assume that the data frame used is already structured in this way. 

## Moose Example Introduction

We also now use the `moose_vignette` data set in the `FPSpatioTemp` package. 

```{r setup}
data(moose_vignette)
```

The rows of `moose_vignette` correspond to sites for moose surveys in the Tok region of Alaska. In general, to use this package, a data frame needs to have, at a bare minimum, four variables:

* two variables of spatial coordinates (an x-coordinate and a y-coordinate)
* one variable of a temporal coordinate (such as day, year, etc.)
* the response variable, with `NA` for any `site-time_point` combination that was not surveyed

The four variables required to perform spatio-temporal finite population prediction in the `moose_vignette` data set are:

* `count`, the number of moose in the site (`NA` if the site was not surveyed)
* `xcoords`, spatial x-coordinates, in Trans-Mercator (TM) units
* `ycoord`, spatial y-coordinates, in TM
* `year`, the year of the survey (either `2019`, `2020`, `2021`, or `2022`).

This data set is a subset of the full Tok data set, which contains more sites and has surveys that date back to 1998. There are 112 sites across 4 years, so the data set has a total of 448 rows.

We will use the `ggplot2` package, the `dplyr` package, and the `tibble` package in the vignette, so we load these packages in with

```{r}
library(ggplot2)
library(dplyr)
library(tibble)
```

```{r}
moose_vignette |> group_by(year) |>
  summarise(n_nonmissing = sum(!is.na(count)))
```

We see that between 21 and 27 sites were surveyed in each year. Again, this is a subset of the larger Tok moose surveys data set; we use this subset so that the models fit more quickly in this vignette. 

## The Basics

In this section, we show how to use the `FPSpatioTemp` package with the basic structure of its two main functions: `stlmfit()` and `predict()`.

We use the `stlmfit()` function to fit a spatio-temporal model to the observed count data. There are five arguments: `formula` and `data` arguments similar to those used in `lm()`, and `xcoord`, `ycoord`, and `tcoord` arguments that give the x-coordinates, y-coordinates, and time coordinates. Note that the model takes a few seconds to fit.

```{r}
mod <- stlmfit(formula = count ~ 1, data = moose_vignette,
               xcoord = "xcoords", ycoord = "ycoords",
               tcoord = "year")
mod
```

The printed output gives any estimated fixed effects coefficients (only an `Intercept` in this example) and a table of estimated covariance parameters. To obtain more information about the fitted model, we can use `summary()`:

```{r}
summary(mod)
```

We can now see the formula used to fit the model, summary statistics on the raw residuals, a table of fixed effects coefficients, standard errors, and p-values, a table of estimated covariance parameters, and the correlation functions used for the spatial and temporal correlation components of the model.

The `plot.stlmfit()` generic function can be used to construct a histogram of the residuals and a residuals vs. fitted values plot. The normality assumption would be important to check if we wanted to obtain a prediction interval for a single site or the total for a small number of sites. However, even with skewed errors, we can still predict the abundance on a large number of sites because the total will still follow a normal distribution from an application of the central limit theorem for dependent data.

```{r}
plot(mod)
```

We then use the `predict.stlmfit()` generic function on the fitted model `mod` to, by default, predict the total abundance of moose in the most recent time point of the survey:

```{r}
predict(mod)
```

The output also gives a standard error for the prediction and the bounds of a 90% prediction interval.

`predict()` also returns a `data` object that has the sitewise predictions for sites with non-zero prediction weights as a column named `predictions_`.

```{r}
predict(mod)$data |> dplyr::filter(year == 2022) |>
  dplyr::relocate(predictions_, everything())
```

The `predictions_` column could then be joined back to a shape file to make a map of the observed survey values with predictions for sites that were not surveyed, as was done in Figure 4 in the manuscript.

#### Other Functionality to `stlmfit()`

The `formula` in `stlmfit()` can be include covariates to be used in the model fitting. For example, in `moose_vignette`, we might consider a model with `strata` as a covariate:

```{r}
mod_strat <- stlmfit(formula = count ~ strata,
                     data = moose_vignette,
                     xcoord = "xcoords", ycoord = "ycoords",
                     tcoord = "year")
mod_strat
```

`mod_strat` could then be passed to `predict()` to obtain a prediction for the total abundance using a model with `strata` as a covariate.

Two other optional arguments to `stlmfit()` include `cor_model_sp` and `cor_model_t`, which control the correlation function to be used for the spatial and temporal model components, respectively. Options for `cor_model_sp` and `cor_model_t` include `"exponential"` (the default for both), `"spherical"`, `"gaussian"`, and `"tent"`, among others. The model below uses `strata` as a covariate, a `"spherical"` correlation structure for the spatial correlation, and a `"spherical"` correlation structure for the temporal correlation.

```{r}
mod_corr <- stlmfit(formula = count ~ strata,
                     data = moose_vignette,
                     xcoord = "xcoords", ycoord = "ycoords",
                     tcoord = "year",
                     cor_model_sp = "spherical",
                     cor_model_t = "spherical")
mod_corr
```

Finally, the `plot_cov()` function can be used on `stlmfit` objects to explore the estimated covariance. Using the original `mod`,

```{r}
plot_cov(mod)
```

By default, the x-axis for temporal distance goes from 0 to the maximum observed temporal distance between two data points (3 years in this case), and the colour for the spatial distance are 0, the minimum observed spatial distance between two data points, and the maximum observed spatial distance between two data points.

The `t_max` and `sp_epstol` arguments to `plot_cov()` can be used to change the maximum temporal distance and provide a vector of spatial distances to use as colours, respectively:

```{r}
plot_cov(mod, sp_epstol = c(0, 4, 40), t_max = 5)
```

#### Other Functionality to `predict()`

The other primary argument of predict is `wts`, which can be used to specify specific weights used for the prediction. By default, `wts` is a vector of `0`'s and `1`'s with a `1` for any observation in the response in the most recent time point and a `0` otherwise. With the default, we are predicting the total abundance in the most recent time point.

`wts` can either be a vector of the same length as the `data` used in `stlmfit()` or `wts` can be a string giving the name of a column in the `data` used in `stlmfit()`. 

To illustrate how `wts` can be used, suppose we are interested in predicting the total abundance across all time points. We can specify the `wts` to be a vector of 1's:

```{r}
total_all_years <- rep(1, nrow(moose_vignette))
predict(mod, wts = total_all_years)
```

The `wts` argument can also be used to incorporate sites with unequal areas. The `area_mi` variable in `moose_vignette` gives the area of each site (in square miles):

```{r}
moose_vignette
```

All areas are roughly equal but there is a little variability in site area. To take into account area, we can fit a model on moose __density__ (count per square mile) and then give `area_mi` as the `wts` argument in `predict()` so that the densities are transformed back into counts. The code below creates a `density` variable to be used as a response and a `area_current` variable to be used in the `wts` argument. `area_current` is equal to the site area of the data point is in the most current time point and is equal to 0 otherwise (so that our prediction will be for the total abundance in the most current time point, not the total abundance across all years).

```{r}
moose_vignette <- moose_vignette |> 
  mutate(density = count / area_mi) |>
  mutate(area_current = if_else(year == max(year),
                                true = area_mi,
                                false = 0))
```

We then fit the model with `density` as the response:

```{r}
mod_density <- stlmfit(formula = density ~ 1,
                       data = moose_vignette,
                       xcoord = "xcoords", ycoord = "ycoords",
                       tcoord = "year")
              
mod_density
```

The `wts` argument can either be a vector, as it was in the previous example, or it can be the name of a column in the data set used in the `data` argument of `stlmfit()`:

```{r}
predict(mod_density, wts = "area_current")
```

Note that the prediction for the total abundance in 2022 incorporating area and not incorporating area are nearly identical because the site areas in the data are nearly identical. 

#### Sampling Frame

In some applications, the sampling frame does not change at all from time point to time point, in which case we do not need to pay as much attention to what sites are included in the data from year to year. However, the moose survey sampling frame for the Tok surveys changed in 2014, when biologists focused more of their survey efforts on a different area (the Taylor Corridor) than they did before 2014. 

If sites that are not in the sampling frame are included in the data set for the most recent time point, then the prediction weights for those sites should be set to `0` in `wts` argument. Otherwise, the `predict()` function will predict total abundance in all sites, both in and out of the sampling frame, for the most recent time point.

#### Forecasting

We may also be interested in using a fitted spatio-temporal model to predict the total abundance in, for example, the time point after the most recent survey. The `moose_vignette` data has `year`s from 2019 through 2022. If we want to predict for the total abundance in 2023, a year for which a survey has not been conducted, we need to append rows with the year `2023` to the `moose_vignette` data. The `count` should be `NA` for these new rows, and there should be one row for each site in the sampling frame.

```{r}
sites_2023 <- moose_vignette |>
  dplyr::filter(year == 2022) |>
  mutate(count = NA, 
         xcoords = xcoords, ycoords = ycoords,
         year = 2023, strata = strata,
         area_mi = area_mi, elev_mean = elev_mean)
sites_2023
```

The `sites_2023` data frame has the same coordinates, strata, area, elevation values, and ID numbers as the sites in the 2022 data. But, the `year` variable is now 2023 and the `count` variable is `NA` for each site.

We can bind the 2023 data to the `moose_vignette` data with

```{r}
moose_vignette_2023 <- bind_rows(moose_vignette, sites_2023)
```

Now, `stlmfit()` and `predict.stlmfit()` can be used to obtain a total abundance prediction for the year 2023:

```{r}
mod_2023 <- stlmfit(count ~ 1, data = moose_vignette_2023,
        xcoord = "xcoords", ycoord = "ycoords",
        tcoord = "year")

predict(mod_2023)
```

The standard error for the total abundance for the forecasting model is higher than the standard error from the previous model because, without survey data from 2023, we are less certain about what the total abundance is.
